---
output: rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, comment="#>")
devtools::load_all(".")
```

# hypeR

[![](https://img.shields.io/badge/bioconductor-3.10-3a6378.svg)](https://doi.org/doi:10.18129/B9.bioc.hypeR)
[![](https://img.shields.io/badge/platforms-linux%20%7C%20osx%20%7C%20win-2a89a1.svg)](https://bioconductor.org/checkResults/3.9/bioc-LATEST/hypeR/)
[![](https://img.shields.io/badge/lifecycle-maturing-4ba598.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![](https://bioconductor.org/shields/build/devel/bioc/hypeR.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/hypeR/)
[![](https://img.shields.io/github/last-commit/montilab/hypeR.svg)](https://github.com/montilab/hypeR/commits/master)

## Documentation

Please visit <https://montilab.github.io/hypeR-docs/>

## Requirements

__hypeR__ currently requires the latest version of R (>= 3.6.0) to be installed directly from Github or Bioconductor. To install with R (>= 3.5.0) see below. Use with R (< 3.5.0) is not recommended.

## Installation

Install the development version of the package from Github.

``` r
devtools::install_github("montilab/hypeR")
```

Or install the development version of the package from Bioconductor.

``` r
BiocManager::install("montilab/hypeR", version="devel")
```

 Or install with  Conda.
```bash
conda create --name hyper
source activate hyper
conda install -c r r-devtools
R
library(devtools)
devtools::install_github("montilab/hypeR")
```

Or install with previous versions of R.
```bash
git clone https://github.com/montilab/hypeR
nano hypeR/DESCRIPTION
# Change Line 8
# Depends: R (>= 3.6.0) -> Depends: R (>= 3.5.0)
R
install.packages("path/to/hypeR", repos=NULL, type="source")
```

## Usage

```{r, eval=FALSE}

library(hypeR)

```

## Terminology
### Signature
__hypeR__ employs multiple types of enrichment analyses (e.g. hypergeometric, kstest, gsea). Depending on the type, different kinds of signatures are expected. There are three types of signatures `hypeR()` expects.
```{r}

# Simply a character vector of symbols (hypergeometric)
signature <- c("GENE1", "GENE2", "GENE3")

# A ranked character vector of symbols (kstest)
ranked.signature <- c("GENE2", "GENE1", "GENE3")

# A ranked named numerical vector of symbols with ranking weights (gsea)
weighted.signature <- c("GENE2"=1.22, "GENE1"=0.94, "GENE3"=0.77)

```

### Geneset
A geneset is simply a list of vectors, therefore, one can use any custom geneset in their analyses, as long as it's appropriately defined.
```{r}

genesets <- list("GSET1" = c("GENE1", "GENE2", "GENE3"),
                 "GSET2" = c("GENE4", "GENE5", "GENE6"),
                 "GSET3" = c("GENE7", "GENE8", "GENE9"))

```

#### Hyper enrichment
All workflows begin with performing hyper enrichment with `hyper()`. Often we are just interested in a single signature, as described above. In this case, `hyper()` will return a `hyp` object. This object contains relevant information to the enrichment results and is recognized by downstream methods.
```{r}

hyp_obj <- hypeR(signature, genesets)

```

#### Downstream methods
Please visit the [documentation](https://montilab.github.io/hypeR-docs/) for detailed functionality. Below is a brief list of some methods.

##### Downloading genesets
```{r, eval=FALSE}

BIOCARTA <- msigdb_gsets(species="Homo sapiens", category="C2", subcategory="CP:BIOCARTA")
KEGG     <- msigdb_gsets(species="Homo sapiens", category="C2", subcategory="CP:KEGG")
REACTOME <- msigdb_gsets(species="Homo sapiens", category="C2", subcategory="CP:REACTOME")

```

##### Visualize results
```{r, eval=FALSE}

# Show interactive table
hyp_show(hyp_obj)

# Plot dots plot
hyp_dots(hyp_obj)

# Plot enrichment map
hyp_emap(hyp_obj)

# Plot hiearchy map
hyp_hmap(hyp_obj)

```

##### Saving results
```{r, eval=FALSE}

# Save to excel
hyp_to_excel(hyp_obj, file_path="hypeR.xlsx")

# Save to table
hyp_to_table(hyp_obj, file_path="hypeR.txt")

# Generate markdown report
hyp_to_rmd(hyp_obj,
           file_path="hypeR.rmd",
           title="Hyper Enrichment (hypeR)",
           subtitle="YAP, TNF, and TAZ Knockout Experiments",
           author="Anthony Federico, Stefano Monti")

```

## Related Repositories

- [hypeR-db](https://github.com/montilab/hypeR-db)
*A repository for commonly used open source genesets used by hypeR*
- [hypeR-shiny](https://github.com/montilab/hypeR-shiny)
*An example shiny application using hypeR*
- [hypeR-docs](https://github.com/montilab/hypeR-docs)
*Landing site for hosting documentation for hypeR*
- [hypeR-workshop](https://github.com/montilab/hypeR-workshop)
*Materials for a hypeR tutorial workshop*

## Cite

Anthony Federico, Stefano Monti (2019) hypeR: An R Package for Geneset Enrichment Workflows. _Bioinformatics_. https://doi.org/10.1093/bioinformatics/btz700
