---
output: rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, comment="#>")
devtools::load_all(".")
```

# hypeR

[![](https://img.shields.io/badge/bioconductor-3.9-3a6378.svg)](https://doi.org/doi:10.18129/B9.bioc.hypeR)
[![](https://img.shields.io/badge/platforms-linux%20%7C%20osx%20%7C%20win-2a89a1.svg)](https://bioconductor.org/checkResults/3.9/bioc-LATEST/hypeR/)
[![](https://img.shields.io/badge/lifecycle-maturing-4ba598.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![](https://bioconductor.org/shields/build/devel/bioc/hypeR.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/hypeR/)
[![](https://img.shields.io/github/last-commit/montilab/hypeR.svg)](https://github.com/montilab/hypeR/commits/master) 

## Documentation

Please visit https://montilab.github.io/hypeR-docs/

## Installation

Install the development version of the package from Github.
```{r, eval=FALSE}

devtools::install_github("montilab/hypeR")

```

Or install the development version of the package from Bioconductor.
```{r, eval=FALSE}

BiocManager::install("montilab/hypeR")

```

If not using the current version of R, try installing from our dev branch.
```{r, eval=FALSE}

devtools::install_github("montilab/hypeR", ref="dev")

```

## Usage

```{r, eval=FALSE}

library(hypeR)

```

## Terminology
### Signature
__hypeR__ employs multiple types of enrichment analyses (e.g. hypergeometric, kstest, gsea). Depending on the type, different kinds of signatures are expected. There are three types of signatures `hypeR()` expects.
```{r}

# Simply a character vector of symbols (hypergeometric)
signature <- c("GENE1", "GENE2", "GENE3")

# A pre-ranked character vector of symbols (kstest)
ranked.signature <-  c("GENE1", "GENE2", "GENE3")

# A pre-ranked named numerical vector of symbols with ranking weights (gsea)
weighted.signature <-  c("GENE1"=1.22, "GENE2"=0.94, "GENE3"=0.77)

```

### Geneset
A geneset is simply a list of vectors, therefore, one can use any custom geneset in their analyses, as long as it's appropriately defined.
```{r}

genesets <- list("GSET1" = c("GENE1", "GENE2", "GENE3"),
                 "GSET2" = c("GENE4", "GENE5", "GENE6"),
                 "GSET3" = c("GENE7", "GENE8", "GENE9"))

```

#### Hyper enrichment
All workflows begin with performing hyper enrichment with `hyper()`. Often we are just interested in a single signature, as described above. In this case, `hyper()` will return a `hyp` object. This object contains relevant information to the enrichment results and is recognized by downstream methods.
```{r}

hyp_obj <- hypeR(signature, genesets)

```

#### Downstream methods
Please visit the [documentation](https://montilab.github.io/hypeR-docs/) for detailed functionality. Below is a brief list of some methods.

##### Downloading genesets
```{r, eval=FALSE}

# Download genesets from msigdb
msigdb_path <- msigdb_download_all(species="Homo sapiens")

BIOCARTA <- msigdb_fetch(msigdb_path, "C2.CP.BIOCARTA")
KEGG     <- msigdb_fetch(msigdb_path, "C2.CP.KEGG")
REACTOME <- msigdb_fetch(msigdb_path, "C2.CP.REACTOME")

```

##### Visualize results
```{r, eval=FALSE}

# Show interactive table
hyp_show(hyp_obj)

# Plot dots plot
hyp_dots(hyp_obj)

# Plot enrichment map
hyp_emap(hyp_obj)

# Plot hiearchy map
hyp_hmap(hyp_obj)

```

##### Saving results
```{r, eval=FALSE}

# Save to excel
hyp_to_excel(hyp_obj, file_path="hyper.xlsx")

# Save to table
hyp_to_table(hyp_obj, file_path="hyper.txt")

# Generate markdown report
hyp_to_rmd(lmultihyp_obj,
           file_path="hyper-enrichment.rmd",
           title="Hyper Enrichment (hypeR)",
           subtitle="YAP, TNF, and TAZ Knockout Experiments",
           author="Anthony Federico, Stefano Monti",
           show_plots=T,
           show_emaps=T,
           show_hmaps=T,
           show_tables=T)

```
