---
title: "Basic Usage"
vignette: >
  %\VignetteIndexEntry{Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#>",
    cache = FALSE)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

### Example of a Signature
Here we define our signature of interest as a group of genes randomly sampled from the *estrogen-dependent gene expression* gset. A signature is simply a vector of symbols.
```{r}

signatures <- readRDS(system.file("extdata/signatures.rds", package="hypeR"))
signature <- signatures$signature

```

```{r}

head(signature)

```

### Example of a Geneset
Here we use an genesets example from *REACTOME*. A geneset is simply a list of vectors, therefore, one can use any custom geneset in their analysis, as long as it is appropriately defined.
```{r}

gsets <- readRDS(system.file("extdata/gsets.rds", package="hypeR"))
gsets <- gsets$REACTOME

```

```{r}

head(names(gsets))
head(gsets[[1]])

```

### Hyper Enrichment
All workflows begin with performing hyper enrichment with __*hypeR*__. Often we are just interested in a single signature, as described above. In this case, __*hypeR*__ will return a *hyp object*. This object contains relevant information to the enrichment results and is recognized by downstream visualization and analysis tools.
```{r}

hyp_obj <- hypeR(signature, gsets, fdr=0.05)

```

```{r}

print(hyp_obj)
hyp_df <- hyp_obj$as.data.frame()
print(hyp_df[5:10,1:4])

```

### Interactive Table
```{r}

hyp_show(hyp_obj, simple=FALSE, stylish=TRUE)

```

### Set Background Population Size
Because genomic experiments often vary in the number of genes probed, it is important to be able to adjust the background population size accordingly to ensure accurate enrichment statistics. 
```{r}

hyp_obj <- hypeR(signature, gsets, bg=10000, fdr=0.05)

```

### Specify a Background Population of Genes
In cases where the background population is small it is advisable to first reduce genesets to their intersection with the background population of genes. By providing a character vector of background genes instead of an integer, hypeR will do just that. Here is an example of an experiment that only uses sex-linked genes and therefore genesets are restricted to only genes included in the background population.

```{r}
url = "https://www.genenames.org/cgi-bin/download/custom?col=gd_app_sym&chr=X&chr=Y&format=text"
r <- httr::GET(url)
text <- httr::content(r, "text", encoding="ISO-8859-1")
text.split <- strsplit(text, "\n")[[1]]
bg <- text.split[2:length(text.split)]
head(bg)
```

```{r}

hyp_obj <- hypeR(signature, gsets, bg=bg, fdr=0.05)

```

### Save Results to Excel
```{r, eval=FALSE}

hyp_to_excel(hyp_obj, file.path="pathways.xlsx")

```

### Save Results to Table
```{r, eval=FALSE}

hyp_to_table(hyp_obj, file.path="pathways.txt")

```