---
title: "Basic Usage"
vignette: >
  %\VignetteIndexEntry{Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="#>", cache=FALSE)
library(dplyr)
library(magrittr)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

### Terminology
#### Signature
__hypeR__ employs multiple types of enrichment analyses (e.g. hypergeometric, kstest, gsea). Depending on the type, different kinds of signatures are expected. There are three types of signatures `hypeR()` expects.
```{r}

# Simply a character vector of symbols (hypergeometric)
signature <- c("GENE1", "GENE2", "GENE3")

# A pre-ranked character vector of symbols (kstest)
ranked.signature <-  c("GENE1", "GENE2", "GENE3")

# A pre-ranked named numerical vector of symbols with ranking weights (gsea)
weighted.signature <-  c("GENE1"=1.22, "GENE2"=0.94, "GENE3"=0.77)

```

#### Geneset
A geneset is simply a list of vectors, therefore, one can use any custom geneset in their analyses, as long as it's appropriately defined. In our tutorials, we will use genesets from [REACTOME](https://reactome.org/). There is also what is called relational genesets, whereby genesets are organized into a hiearchy; we will explore these in later tutorials. 
```{r}

genesets <- list("GSET1" = c("GENE1", "GENE2", "GENE3"),
                 "GSET2" = c("GENE4", "GENE5", "GENE6"),
                 "GSET3" = c("GENE7", "GENE8", "GENE9"))

```

### Example Data
In these tutorials, we will use example data. The example data includes an expression set object as well as pre-computed results from common workflows such as diffential expression and weighted gene co-expression analyses. 

```{r}

data(hypdat)

```

#### Defining Signatures and Genesets
Using a differential expression dataframe created with limma, we will extract a signature of upregulated genes for use with a hypergeometric test and rank genes descending by their differential expression level for use with a kstest. We'll also import a genesets from [REACTOME](https://reactome.org/).
```{r}

limma <- hypdat$limma

signature <- limma %>% filter(t > 0 & fdr < 0.001) %>% use_series(symbol)

ranked.signature <- limma %>% arrange(desc(t)) %>% use_series(symbol)

gsets <- hypdat$gsets

```

### Basic Usage
All workflows begin with performing hyper enrichment with `hypeR()`. Often we're just interested in a single signature, as described above. In this case, `hypeR()` will return a `hyp` object. This object contains relevant information to the enrichment results and is recognized by downstream methods.
#### Overrepresentation analysis
```{r}

hyp_obj <- hypeR(signature, gsets, test="hypergeometric", bg=50000, fdr_cutoff=0.01, do_plots=TRUE)

```

```{r}

print(hyp_obj)
hyp_df <- hyp_obj$as.data.frame()
print(head(hyp_df[,1:3]), row.names=FALSE)
hyp_obj$plots[[1]]

```

#### Enrichment analysis
```{r}

hyp_obj <- hypeR(ranked.signature, gsets, test="kstest", fdr_cutoff=0.01, do_plots=TRUE)

```

```{r}

print(hyp_obj)
hyp_df <- hyp_obj$as.data.frame()
print(head(hyp_df[,1:3]), row.names=FALSE)
hyp_obj$plots[[1]]

```

### Save results to excel
```{r, eval=FALSE}

hyp_to_excel(hyp_obj, file_path="hyper.xlsx")

```

### Save results to table
```{r, eval=FALSE}

hyp_to_table(hyp_obj, file_path="hyper.txt")

```
