---
title: "Multiple Signatures"
vignette: >
  %\VignetteIndexEntry{Multiple Signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#>",
    fig.align="center",
    fig.width=8.75,
    cache = FALSE)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

### Multiple Signatures
Other times we are looking at multiple signatures that may have been defined in a single experiment. In this case, we can feed __*hypeR*__ a named list of signatures and a *multihyp object* will be returned. This object is essentially just multiple *hyp objects*. However it is recognized and handled differently by downstream visualization and analysis tools.
```{r}

signatures <- readRDS(system.file("extdata/signatures.rds", package="hypeR"))
experiment <- signatures$experiment

```

```{r}

names(experiment)
head(experiment[[1]])

```

### Example of a Geneset
Here we use an genesets example from *REACTOME*.
```{r}

gsets <- readRDS(system.file("extdata/gsets.rds", package="hypeR"))
gsets <- gsets$REACTOME

```

### Hyper Enrichment
All workflows begin with performing hyper enrichment with __*hypeR*__.
```{r}

multihyp_obj <- hypeR(experiment, gsets, bg=35423)

```

```{r}

print(multihyp_obj)

```

### Save Results to Excel
When saving a *multihyp object* to excel, each signature is exported to its own tab.
```{r, eval=FALSE}

hyp_to_excel(multihyp_obj, file.path="pathways.xlsx")

```

### Save Results to Table
When saving a *multihyp object* to a table, each signature is exported as its own table in a single directory.
```{r, eval=FALSE}

hyp_to_table(multihyp_obj, file.path="pathways.txt")

```

#### Plot Results
When plotting a *multihyp object*, each signatures is plotted and returned separately.
```{r}

p <- hyp_plot(multihyp_obj, return_plots=TRUE)

```

```{r}

names(p)

```

```{r fig.height=5}

p[[1]]

```

#### Enrichment Map
When plotting a *multihyp object*, each signatures is plotted and returned separately.
```{r}

p <- hyp_emap(multihyp_obj, top=100, similarity_cutoff=0.2, val="pval", multihyp_titles=FALSE, return_plots=TRUE)

```

```{r}

names(p)

```

```{r fig.height=7}

p[[1]]

```


