---
title: "Multiple Signatures"
vignette: >
  %\VignetteIndexEntry{Multiple Signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="#>", fig.align="center", fig.width=8.75, cache=FALSE)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

### Multiple signatures
Other times we are looking at multiple signatures that may have been defined in a single experiment. In this case, we can feed `hypeR()` a named list of signatures and a `multihyp` object will be returned. This object is essentially just multiple `hyp` objects. However it is recognized and handled differently by downstream visualization and analysis tools.
```{r}

data(hypdat)
wgcna <- hypdat$wgcna
gsets <- hypdat$gsets

```

```{r}

signatures <- wgcna[[1]] 
head(names(signatures))
head(signatures[[1]])

```

### Hyper enrichment
All workflows begin with performing hyper enrichment with `hypeR()`.
```{r}

multihyp_obj <- hypeR(signatures, gsets, test="hypergeometric", bg=50000, fdr_cutoff=0.01, do_plots=FALSE)

```

### Save results to excel
When saving a `multihyp` object to excel, each signature is exported to its own tab.
```{r, eval=FALSE}

hyp_to_excel(multihyp_obj, file_path="pathways.xlsx")

```

### Save results to table
When saving a `multihyp` object to a table, each signature is exported as its own table in a single directory.
```{r, eval=FALSE}

hyp_to_table(multihyp_obj, file_path="pathways.txt")

```

### Visualizations

When plotting a `multihyp` object, each signatures is plotted and returned separately. The same goes for `hyp_dots()`, `hyp_emap()`, and `hyp_hmap()`. 

```{r}

p <- hyp_dots(multihyp_obj, show_plots=FALSE, return_plots=TRUE)

```

```{r}

names(p)[1:3]

```

```{r, fig.height=5}

p[1:3]

```
