---
title: "Relational Genesets"
vignette: >
  %\VignetteIndexEntry{Relational Genesets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(
    comment = "#>",
    fig.align="center",
    fig.width=8.75,
    cache = FALSE)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

#### Example signature
Here we define our signature of interest as a group of genes randomly sampled from the *estrogen-dependent gene expression* geneset.
```{r}

signatures <- readRDS(system.file("extdata/signatures.rds", package="hypeR"))
signature <- signatures$signature

```

#### Relational genesets
When dealing with hundreds of genesets, it's often useful to understand the relationships between them. This allows researchers to summarize many enriched pathways as more general biological processes. To do this, we rely on curated relationships defined between them. For example, [REACTOME](https://reactome.org/) conveniently defines their genesets in a [hiearchy of pathways](https://reactome.org/PathwayBrowser/). This data can be formatted into a relational genesets object called __*rgsets*__.

```{r}

rgsets_lst <- readRDS(system.file("extdata/rgsets.rds", package="hypeR"))
rgsets_obj <- rgsets_lst$REACTOME

```

Relational genesets have three data atrributes including gsets, nodes, and edges. Using hierarchical terms, the *gsets* attribute includes the geneset information for the *leaf nodes* of the hiearchy, the *nodes* attribute describes all nodes in the hiearchical, including *internal nodes*, and the *edges* attribute describes the edges in the hiearchy.

#### $gsets
```{r}

gsets <- rgsets_obj$gsets
names(gsets)[800:805]

```

#### $nodes
```{r}

nodes <- rgsets_obj$nodes
nodes[1123:1128,]

```

#### $edges
```{r}

edges <- rgsets_obj$edges
edges[1994:1999,]

```

#### Hyper enrichment
All workflows begin with performing hyper enrichment with __*hypeR*__. Note that when passing an __*rgsets*__ object, ensure `gsets_relational=TRUE`. Otherwise, __*hypeR*__ will treat it as a list of unstructured gsets. 
```{r}

hyp_obj <- hypeR(signature, rgsets_obj, gsets_relational=TRUE, bg=7842, fdr=0.05)

```

#### Hiearchy map
One can visualize the top enriched pathways using __*hyp_hmap*__ which will return a hiearchy map. Each node represents a geneset, where the shade of the gold border indicates the normalized significance of enrichment. Hover over the *leaf nodes* to view the raw value. Double click *internal nodes* to cluster their first degree connections.s Edges represent a directed relationship between genesets in the hiearchy.
```{r fig.height=7}

hyp_hmap(hyp_obj, top=60, return_plots=TRUE)

```
