---
title: "Generating Reports"
vignette: >
  %\VignetteIndexEntry{Generating Reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#>",
    fig.align="center",
    fig.width=8.75,
    cache = FALSE)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

### Multiple Experiments
Sometimes, we are generating reports for mutliple signatures across multiple experiments. Typically, the best way to manage this data is to process it all at once and generate a markdown report to analyze ourselves or share with our collaborators.

```{r}

signatures <- readRDS(system.file("extdata/signatures.rds", package="hypeR"))
project <- signatures$project

```

```{r}

names(project)
names(project[[1]])
head(project[[1]][[1]])

```

### Example of a Geneset
Here we use an genesets example from *REACTOME*.
```{r}

gsets <- readRDS(system.file("extdata/gsets.rds", package="hypeR"))
gsets <- gsets$REACTOME

```

### Hyper Enrichment
All workflows begin with performing hyper enrichment with __*hypeR*__.
```{r}

lmultihyp_obj <- lapply(project, function(x) hypeR(x, gsets, bg=35423))

```

```{r}

names(lmultihyp_obj)
print(lmultihyp_obj[[1]])

```

#### Markdown Reports
```{r eval=FALSE}
hypeR::hyp_to_rmd(lmultihyp_obj,
                  file_path="hyper-enrichment.rmd",
                  title="Hyper Enrichment (hypeR)",
                  subtitle="YAP, TNF, and TAZ Knockout Experiments",
                  author="Anthony Federico, Stefano Monti",
                  show_plots=T,
                  show_emaps=T,
                  show_tables=T)
```
