---
title: "Downloading Data"
vignette: >
  %\VignetteIndexEntry{Downloading Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="#>", cache=FALSE)
devtools::load_all(".")
```

### Downloading genesets from msigdb
For most purposes, the genesets hosted by [msigdb](https://software.broadinstitute.org/gsea/msigdb/collections.jsp) are more than adequate to perform geneset enrichment analysis. There are various types of genesets available across many species. Therefore, we have added convenient functions for retrieving msigdb data compatible with __hypeR__.

#### Available genesets
```{r}

msigdb_info()

```

#### Downloading and loading genesets
Use `msigdb_download_one()` to download a single geneset into memory.
```{r}

HALLMARK <- msigdb_download_one(species="Homo sapiens", category="H")

```

```{r}

head(names(HALLMARK))
head(HALLMARK[[1]])

```

Use `msigdb_download_all()` to retrieve all genesets for a given species. By default, genesets are cached in a temporary directory, therefore a path object is returned. Use the path object with `msigdb_fetch()` to load genesets into memory. Users can also specify a directory to download to for repeated use.
```{r}

msigdb_path <- msigdb_download_all(species="Homo sapiens")

BIOCARTA <- msigdb_fetch(msigdb_path, "C2.CP.BIOCARTA")
KEGG     <- msigdb_fetch(msigdb_path, "C2.CP.KEGG")
REACTOME <- msigdb_fetch(msigdb_path, "C2.CP.REACTOME")

```

In this example, we are interested in all three of the following genesets, therefore we concatenate them. A geneset is simply a named list of vectors, therefore, one can use any custom genesets in their analysis, as long as it's appropriately defined.
```{r}

gsets <- c(BIOCARTA, KEGG, REACTOME)

```

### Defining custom genesets
As mentioned previously, one can use custom genesets with __hypeR__. In this example, we download one of the many publicly available genesets hosted by [Enrichr](https://amp.pharm.mssm.edu/Enrichr/). Once downloaded, one performs enrichment as normal.
```{r, eval=FALSE}

url = "http://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=Cancer_Cell_Line_Encyclopedia"
r <- httr::GET(url)
text <- httr::content(r, "text", encoding="ISO-8859-1")
text.split <- strsplit(text, "\n")[[1]]
gsets <- sapply(text.split, function(x) {
    genes <- strsplit(x, "\t")[[1]]
    return(genes[3:length(genes)])
})
names(gsets) <- unlist(lapply(text.split, function(x) strsplit(x, "\t")[[1]][1]))

```

### Downloading genesets from hyperdb
If msigdb genesets are not sufficient, we have also provided another set of functions for downloading and loading other open source genesets. Essentially, we wrap the method used in the example defining custom genesets. This is facilitated by interfacing with the publicly available [libraries](https://amp.pharm.mssm.edu/Enrichr/#stats) hosted by [Enrichr](https://amp.pharm.mssm.edu/Enrichr/).

#### Downloading and loading genesets
```{r, eval=FALSE}

gsets <- hyperdb_fetch(type="gsets", "Cancer_Cell_Line_Encyclopedia")

```

#### Available genesets
```{r}

hyperdb_info(type="gsets", quiet=TRUE)[1:15]

```

#### Available relational genesets
```{r}

hyperdb_info(type="rgsets", quiet=TRUE)

```