---
title: "Extending hypeR"
vignette: >
  %\VignetteIndexEntry{Visualize Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="#>", fig.align="center", fig.width=8.75, cache=FALSE)
library(dplyr)
library(magrittr)
library(fgsea)
devtools::load_all(".")
```

hypeR is an R package focused on the downstream analysis of geneset enrichment workflows. Because there are many upstream tools for performing enrichment, users may opt for those while also planning to use the visualization, summarization, and reporting features offered by hypeR. This task is straightforward given the object oriented nature of hypeR.

In this example, we will perform enrichment on a differential expression signature using **fgsea**, a popular R package for Fast Gene Set Enrichment Analysis, and coerce the results into a `hyp` object, compatible with all hypeR methods.

### Load packages
```{r, eval=FALSE}

library(dplyr)
library(magrittr)
library(fgsea)
library(hypeR)

```

### Signature
```{r}
hypdat <- readRDS(file.path(system.file("extdata", package="hypeR"), "hypdat.rds"))

# Differential expression results
df <- hypdat$limma

print(df[1:5,], row.names=FALSE)

# Rank genes by upregulation
df.up <- df %>%
         filter(t > 0 & fdr < 0.001) %>%
         arrange(desc(t))

print(df.up[1:5,], row.names=FALSE)

# Extract a weighted signature for upregulation
signature <- df$t

names(signature) <- df$symbol
head(signature)

```

### Relational genesets
```{r}

rgsets <- hyperdb_fetch(type="rgsets", "REACTOME")

```

### Wrapping *fgsea* to return a `hyp` object
```{r}

fgsea_to_hyp <- function(signature,
                         gsets, 
                         nperm=10000, 
                         direction=c("up", "dn"),
                         pval_cutoff=1, 
                         fdr_cutoff=1,
                         ...) {
    
    # Default arguments
    direction <- match.arg(direction)

    # Save original arguments
    args <- as.list(environment())
    
    # Check if gsets are relational
    if (is(gsets, "rgsets")) gsets <- gsets$gsets

    # Handling of multiple signatures
    if (is(signature, "list")) {
        lhyp <- lapply(signature, fgsea_to_hyp, args$gsets,
                                                args$nperm,
                                                args$direction,
                                                args$pval_cutoff,
                                                args$fdr_cutoff, 
                                                ...)
 
        # Wrap list of hyp objects in multihyp object
        multihyp.obj <- multihyp$new(data=lhyp)
        return(multihyp.obj)
    }

    data <- data.frame(matrix(ncol=6, nrow=0))
    colnames(data) <- c("label", "pval", "fdr", "gset.size", "genes.found", "score")
    plots <- ggempty()    

    # Run fgsea
    results <- fgsea(stats=signature, pathways=gsets, nperm=nperm, ...)
    
    # Reformat results
    results <- data.frame(results)
    colnames(results) <- c("label", "pval", "fdr", "score", "nes", "nme", "genes.found", "le")
    results$pval <- signif(results$pval, 2)
    results$fdr <- signif(results$fdr, 2)
    results$gset.size <- sapply(results$label, function(x) length(gsets[[x]]))
    results <- results[,c("label", "pval", "fdr", "gset.size", "genes.found", "score")]
    
    #If hits are found format dataframe
    if (!is.null(results)) {
        data <- results %>%
                dplyr::filter(pval <= pval_cutoff) %>%
                dplyr::filter(fdr <= fdr_cutoff) %>%
                dplyr::arrange(pval)
        
        if (direction == "up") data %>% dplyr::filter(score > 0)
        if (direction == "down") data %>% dplyr::filter(score < 0)
        
        # Empty plots
        plots <- setNames(data$label, data$label) %>%
                 lapply(function(x) ggempty())
    }

    # Wrap dataframe in hyp object
    hyp_obj <- hyp$new(data=data, plots=plots, args=args)
    return(hyp_obj)
}

hyp_obj <- fgsea_to_hyp(signature, rgsets)

```

### *hypeR* functionality
```{r}

print(hyp_obj)

```

```{r}

hyp_dots(hyp_obj, top=20, show_plots=FALSE, return_plots=TRUE)

```

```{r}

hyp_emap(hyp_obj, top=100, similarity_cutoff=0.7, show_plots=FALSE, return_plots=TRUE)

```

```{r}

hyp_hmap(hyp_obj, top=50, val="fdr", show_plots=FALSE, return_plots=TRUE)

```

```{r, eval=FALSE}

hyp_to_table(hyp_obj, file_path="fsgea.txt")

```

```{r, eval=FALSE}

hyp_to_excel(hyp_obj, file_path="fsgea.xlsx")

```

```{r, eval=FALSE}

hyp_to_rmd(hyp_obj, file_path="fgsea.Rmd")

```
