---
title: "Generating Reports"
vignette: >
  %\VignetteIndexEntry{Generating Reports}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="", fig.align="center", fig.width=8.75, cache=FALSE)
devtools::load_all(".")
```

Often researchers need to annotate multiple signatures across multiple experiments. Typically, the best way to manage this data is to process it all at once and generate a markdown report to analyze and share with collaborators. Let's take an example where we have one experiment with many signatures.

```{r}
data(wgcna)

signatures <- wgcna[[1]]
str(signatures)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
genesets <- hyperdb_rgsets("KEGG", "92.0")
mhyp <- hypeR(signatures, genesets, test="hypergeometric", background=36000, fdr=0.05)
```

We have included some convenient functions that return some html for turning `hyp` and `mhyp` objects into concise tables for markdown documents. These tables can also include some plots for each individual `hyp` object.
```{r}
rctbl_mhyp(mhyp, show_hmaps=TRUE)
```

Alright, how about we process all of these nested signatures and provide an example of what your markdown report might look like. Because we have a nested list of signatures, we use `hyper()` with the `lapply` function. This results in a list of `mhyp` objects.

```{r, message=FALSE, warning=FALSE, results='hide'}
lmhyp <- lapply(wgcna, hypeR, genesets, test="hypergeometric", background=36000, fdr=0.05)
```

First start with a nice theme and include the css required to make the tables pretty.

```r
title: "Project Title"
subtitle: "Project Subtitle"
author: "Anthony Federico"
date: "Last Modified: 'r Sys.Date()'"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: false
    code_folding: hide
    toc_depth: 4
    css: 'r system.file("style", "rctbl.css", package="hypeR")'
```


### LA

```{r}
hyp_dots(lmhyp$LA, merge=TRUE)
rctbl_mhyp(lmhyp$LA, show_hmaps=TRUE)
```

### LB
```{r}
hyp_dots(lmhyp$LB, merge=TRUE)
rctbl_mhyp(lmhyp$LB, show_hmaps=TRUE)
```

### H2
```{r}
hyp_dots(lmhyp$H2, merge=TRUE)
rctbl_mhyp(lmhyp$H2, show_hmaps=TRUE)
```

### BL
```{r}
hyp_dots(lmhyp$BL, merge=TRUE)
rctbl_mhyp(lmhyp$BL, show_hmaps=TRUE)
```

### Session Info
```{r}
sessionInfo()
```

### Generating Markdown Reports

For quick report generation one can use `hyp_to_rmd()` which will accept multiple formats, including a single `hyp` or `multihyp` object as well as a list of either, including a list of `hyp` or `multihyp` objects together. When a list of `multihyp` objects are passed for example, each experiment will become its own section, while each signature becomes its own tab within that section. Lists of keyword arguments can be passed for `hyp_dots()`, `hyp_emap()`, and `hyp_hmap()`, allowing customization of their functionality per report. See `hyp_to_rmd()` for details.

```{r, eval=FALSE}
hyp_to_rmd(lmultihyp_obj,
           file_path="hypeR.rmd",
           title="hypeR Report",
           subtitle="Weighted Gene Co-expression Analysis",
           author="Anthony Federico, Stefano Monti",
           show_dots=T,
           show_emaps=T,
           show_hmaps=T,
           show_tables=T)
```
