---
title: "Generating Reports"
vignette: >
  %\VignetteIndexEntry{Generating Reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="", fig.align="center", fig.width=8.75, cache=FALSE)
devtools::load_all(".")
```

### Multiple Experiments
Often researchers need to annotate multiple signatures across multiple experiments. Typically, the best way to manage this data is to process it all at once and generate a markdown report to analyze and share with collaborators.

```{r}

hypdat <- readRDS(file.path(system.file("extdata", package="hypeR"), "hypdat.rds"))

```

In the following example, we're interested in exploring the signatures from a [Weighted Gene Co-expression Analysis](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/) of two different phenotypes. In `hypdat`, we take the resulting co-expression modules from WGCNA ran separately on gene expression data from carcinogenic and non-carcinogenic compounds. There are 15 modules of co-expressed genes for carcinogenic compounds and 16 modules of co-expressed genes for non-carcinogenic compounds. Using `hyp_to_rmd()` will help us process this data in a concise and reproducible manner that is readily shareable.

```{r}

signatures <- hypdat$wgcna
str(signatures)

```

Because we have a nested list of signatures, we use `hyper()` with the `lapply` function. This results in a list of `multihyp` objects, which can be directly passed to `hyp_to_rmd()`.

```{r, eval=FALSE}

genesets <- hyperdb_rgsets("REACTOME", "70.0")

lmultihyp_obj <- lapply(signatures, hypeR, genesets, test="hypergeometric", background=50000, fdr=0.01)

```

### Generating Markdown Reports

`hyp_to_rmd()` will accept multiple formats, including a single `hyp` or `multihyp` object as well as a list of either, including a list of `hyp` or `multihyp` objects together. 

When a list of `multihyp` objects are passed for example, each experiment will become its own section, while each signature becomes its own tab within that section. 

Lists of keyword arguments can be passed for `hyp_dots()`, `hyp_emap()`, and `hyp_hmap()`, allowing customization of their functionality per report. See `hyp_to_rmd()` for details.

```{r, eval=FALSE}

hyp_to_rmd(lmultihyp_obj,
           file_path="hypeR.rmd",
           title="hypeR Report",
           subtitle="Weighted Gene Co-expression Analysis",
           author="Anthony Federico, Stefano Monti",
           show_dots=T,
           show_emaps=T,
           show_hmaps=T,
           show_tables=T)

```

An example of these types of reports can be viewed at https://montilab.github.io/hypeR-reporting.
