---
title: "Generating Reports"
vignette: >
  %\VignetteIndexEntry{Generating Reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="#>", fig.align="center", fig.width=8.75, cache=FALSE)
devtools::load_all("/Users/anthonyfederico/GitHub/hypeR")
```

### Multiple experiments
Sometimes, we are generating reports for multiple signatures across multiple experiments. Typically, the best way to manage this data is to process it all at once and generate a markdown report to analyze ourselves or share with collaborators.

In the following example, we're interested in exploring the signatures from a weighted gene co-expression analysis (WGCNA) of two different phenotypes. In `hypdat`, we take the resulting co-expression modules from WGCNA ran separately on gene expression data from carcinogenic and non-carcinogenic compounds. There are 15 modules of co-expressed genes for carcinogenic compounds and 16 modules of co-expressed genes for non-carcinogenic compounds. The `hyp_to_rmd()` will help us process this data in a concise and reproducible manner that is readily shareable.
```{r}

data(hypdat)
wgcna <- hypdat$wgcna
names(wgcna)
names(wgcna[["NON-CARC"]])
names(wgcna[["CARCINOGEN"]])

rgsets <- hypdat$rgsets

```

### Hyper enrichment
All workflows begin with performing hyper enrichment with `hypeR()`.
```{r, eval=FALSE}

lmultihyp_obj <- lapply(wgcna, hypeR, rgsets, test="hypergeometric", bg=50000, fdr_cutoff=0.01, do_plots=FALSE)

```

### Markdown reports
`hyp_to_rmd()` will accept multiple formats, including a single `hyp` or `multihyp` object as well as a list of either, including a list of `hyp` or `multihyp` objects together. When a list of `multihyp` objects are passed for example, each experiment will become its own section, while each signature becomes its own tab within that section. Lists of keyword arguments can be passed for `hyp_dots()`, `hyp_emap()`, and `hyp_hmap()`, allowing customization of their functionality per report.
```{r, eval=FALSE}

hyp_to_rmd(lmultihyp_obj,
           file_path="hyper-enrichment.rmd",
           title="Hyper Enrichment (hypeR)",
           subtitle="Weighted Gene Co-expression Analysis",
           author="Anthony Federico, Stefano Monti",
           show_dots=F,
           show_emaps=F,
           show_hmaps=T,
           show_tables=T)

```
