---
title: "Constructing Relational Genesets"
vignette: >
  %\VignetteIndexEntry{Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="#>", cache=FALSE)
library(dplyr)
library(qusage)
library(igraph)
library(magrittr)
devtools::load_all(".")
```

### Relational genesets

When dealing with hundreds of genesets, it's often useful to understand the relationships between them. This allows researchers to summarize many enriched pathways as more general biological processes. To do this, we rely on curated relationships defined between them. For example, [REACTOME](https://reactome.org/) conveniently defines their genesets in a [hiearchy of pathways](https://reactome.org/PathwayBrowser/). This data can be formatted into a relational genesets object called `rgsets`.

#### Data Sources
```{r}

# All data downloaded from https://reactome.org/download-data
gsets.url <- "https://reactome.org/download/current/ReactomePathways.gmt.zip"
nodes.url <- "https://reactome.org/download/current/ReactomePathways.txt"
edges.url <- "https://reactome.org/download/current/ReactomePathwaysRelation.txt"

```

#### Loading Data
```{r}
# Gsets
gsets.tmp <- tempfile(fileext=".gmt.zip")
download.file(gsets.url, destfile = gsets.tmp, mode = "wb")
gsets.raw <- gsets.tmp %>%
             unzip() %>%
             read.gmt() %>%
             lapply(function(x) {
                 toupper(x[x != "Reactome Pathway"])
             })
# Nodes
nodes.raw <- nodes.url %>%
             read.delim(sep="\t", 
                        header=FALSE, 
                        fill=TRUE, 
                        col.names=c("id", "label", "species"), 
                        stringsAsFactors=FALSE)
# Edges
edges.raw <- edges.url %>%
             read.delim(sep="\t", 
                        header=FALSE, 
                        fill=TRUE, 
                        col.names=c("from", "to"),
                        stringsAsFactors=FALSE)
```

#### Organizing a Hierarchy
```{r}
# Species-specific nodes
nodes <- nodes.raw %>%
         filter( label %in% names(gsets.raw) ) %>%
         filter( species == "Homo sapiens" ) %>%
         filter(! duplicated(id) ) %>%
         set_rownames( .$id ) %>%
         { .[, "label", drop=FALSE] }

# Species-specific edges
edges <- edges.raw %>%
         filter( from %in% rownames(nodes) ) %>%
         filter( to %in% rownames(nodes) )

# Leaf genesets
gsets <- nodes %>%
         rownames() %>%
         .[! . %in% edges$from] %>%
         sapply( function(x) nodes[x, "label"] ) %>%
         gsets.raw[.]
```

#### Using the `rgsets` object
```{r}
rgsets <- hypeR::rgsets$new(gsets, nodes, edges)
```

#### Visualize
```{r}
par(mar=c(0,0,0,0))
g <- graph_from_edgelist(as.matrix(rgsets$edges))
plot(g, vertex.label=NA, vertex.size=1, edge.arrow.size=0)
```