---
title: "Constructing Relational Genesets"
vignette: >
  %\VignetteIndexEntry{Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="", fig.align="center", cache=FALSE)
library(dplyr)
library(qusage)
library(igraph)
library(magrittr)
devtools::load_all(".")
```

### Relational genesets

When dealing with hundreds of genesets, it's often useful to understand the relationships between them. This allows researchers to summarize many enriched pathways as more general biological processes. To do this, we rely on curated relationships defined between them. For example, [REACTOME](https://reactome.org/) conveniently defines their genesets in a [hiearchy of pathways](https://reactome.org/PathwayBrowser/). This data can be formatted into a relational genesets object called `rgsets`.

#### Data sources
Raw data for gsets, nodes, and edges can be directly downloaded.
```{r}

# All data downloaded from https://reactome.org/download-data
gsets.url <- "https://reactome.org/download/current/ReactomePathways.gmt.zip"
nodes.url <- "https://reactome.org/download/current/ReactomePathways.txt"
edges.url <- "https://reactome.org/download/current/ReactomePathwaysRelation.txt"

```

#### Loading data
```{r}
# Gsets
gsets.tmp <- tempfile(fileext=".gmt.zip")
download.file(gsets.url, destfile = gsets.tmp, mode = "wb")
gsets.raw <- gsets.tmp %>%
             unzip() %>%
             read.gmt() %>%
             lapply(function(x) {
                 toupper(x[x != "Reactome Pathway"])
             })
# Nodes
nodes.raw <- nodes.url %>%
             read.delim(sep="\t", 
                        header=FALSE, 
                        fill=TRUE, 
                        col.names=c("id", "label", "species"), 
                        stringsAsFactors=FALSE)
# Edges
edges.raw <- edges.url %>%
             read.delim(sep="\t", 
                        header=FALSE, 
                        fill=TRUE, 
                        col.names=c("from", "to"),
                        stringsAsFactors=FALSE)
```

#### Organizing a hierarchy
```{r}
# Species-specific nodes
nodes <- nodes.raw %>%
         filter( label %in% names(gsets.raw) ) %>%
         filter( species == "Homo sapiens" ) %>%
         filter(! duplicated(id) ) %>%
         set_rownames( .$id ) %>%
         { .[, "label", drop=FALSE] }

# Species-specific edges
edges <- edges.raw %>%
         filter( from %in% rownames(nodes) ) %>%
         filter( to %in% rownames(nodes) )

# Leaf genesets
gsets <- nodes %>%
         rownames() %>%
         .[! . %in% edges$from] %>%
         sapply( function(x) nodes[x, "label"] ) %>%
         gsets.raw[.]
```

##### nodes
A single-column data frame of labels where the rownames are unique identifiers. Leaf node labels should have an associated geneset, while internal nodes do not have to. The only genesets tested, will be those in the list of genesets.
```{r}
head(nodes)
```

##### edges
A dataframe with two columns of identifiers, indicating directed edges between nodes in the hierarchy.
```{r}
head(edges)
```

##### gsets
A list of character vectors, named by the geneset labels. Typically, genesets will be at the leaves of the hierarchy, while not required.
```{r}
head(names(gsets))
```

#### Using the `rgsets` object
```{r}
rgsets <- hypeR::rgsets$new(gsets, nodes, edges)
```

#### Visualize
```{r}
par(mar=c(0,0,0,0))
g <- graph_from_edgelist(as.matrix(rgsets$edges))
plot(g, vertex.label=NA, vertex.size=1, edge.arrow.size=0)
```